% !TEX root = dynamicslearning.tex
\section{Introduction}

What are the instinctive individual reactions which make a group of animals forming coordinated movements, for instance a flock of migrating birds or a school of fish? Which biochemical interactions between cells produce the formation of complex structures, like tissues and organs? What are the mechanisms, which induce certain significant changes in a large amount of players in the financial market? 
In this paper we are concerned with the ``mathematization'' of the problem of learning or inferring interaction rules from observations of evolutions. The framework we consider is the one of  evolutions driven by gradient descents.
The study of gradient flow evolutions to minimize  certain energetic landscapes has been the subject of intensive research in the past years \cite{AGS}. Some of the most recent models are  aiming at describing time-dependent phenomena also in biology or even in social dynamics, borrowing a leaf from more established and classical  models in physics. 

{
 For instance, starting with the seminal papers of Vicsek et. al. \cite{VCBCS95} and Cucker-Smale \cite{CucSma07}, there has been a flood of models describing collective motion, based on the exchange of information as long-range social interactions (forces) between active agents (particles), %see for example \cite{parisi08,parisi08-1,parisi08-2,Hildenbrandt01112010,mann11,heoemascszwa11,Motsch2011} 
see for example \cite{cafotove10} and references therein. Besides these very relevant second order models, very popular to describe collective motion, also first order systems have often been used to model multiagent interactions, e.g., in opinion formation \cite{kr00,hekr02}, vehicular traffic flow \cite{garavello2006traffic}, pedestrian motion \cite{crpito14}, and synchronisation of chemical and biological oscillators in neuroscience \cite{kuramoto2003chemical}, to mention a few\footnote{The literature on the subjects above is by now so vast that it is impossible to give a complete account of it, and we limit ourselves to mention some seminal or introductory references.}.}


However, for the analysis, but even more crucially for the reliable and realistic numerical simulation of such phenomena, one presupposes a complete understanding and determination of the governing { interactions}. Unfortunately, except for physical situations where the calibration of the model can be done by measuring the governing forces rather precisely, for some relevant macroscopical models in physics and most of the models in biology and social sciences the governing interactions are far from being precisely determined. In fact, very often in these studies the governing potential energies are just predetermined to be able to reproduce, at least approximately or qualitatively, some of the macroscopical effects of the observed dynamics, such as the formation of certain patterns, but there has been relatively little effort in the applied mathematics literature towards matching data from real-life cases. 

%This attitude aiming just at a qualitative description tends however to reduce  some of the investigations in this area to  %beautiful and mathematically interesting toy-cases, which have likely little to do with real-life scenarios. 
In this paper we aim at bridging, in the specific setting of {deterministic} first order models, the well-developed theory of dynamical systems and mean-field equations with classical  approaches of approximation theory, nonlinear time series analysis, and machine learning. We  provide a  mathematical framework for the reliable identification of the governing forces from data obtained by direct observations of corresponding time-dependent evolutions. This is a new kind of inverse problem, beyond more traditionally considered ones, as the forward map is a strongly nonlinear  evolution, highly dependent on the probability measure generating the initial conditions.  {Below we also advance reasons for classical optimal likely to  fail providing reliable methods of dynamics learning in this context.} As we aim at a precise quantitative analysis, and to be very concrete, we will  attack the learning of the governing laws of evolution for specific models in social dynamics governed by nonlocal interactions.  { Motivated by the vast literature mentioned above, the models considered in the scope of this paper are deterministic and of  first order,  however we intend in follow up work to extend our results towards stochastic dynamical systems and higher order dynamics.}


